---
title: "Functions in R"
author: "Mburu"
date: '2022-07-01'
format:
  revealjs:
     theme: sky
     highlight: pygments
     transition: convex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)


library(tidyverse)
library(data.table)
library(DT)
```



## Functions

- A function is a block of organized, reusable code that is used to perform a single, related action. Functions provide better modularity for your application and a high degree of code reusing

-----

### Functions Already used

```{r}
x <- 1:10
y <- 11:20
sum(x)
mean(x)
cor(x, y)
```

-----

```{r}
lm_fit <- lm(Petal.Length ~ Petal.Width, data = iris)

library(broom)


```

------

```{r}

cols <- c( "estimate", "std.error", "statistic", "p.value")
tidy_lm <- tidy(lm_fit) 
tidy_lm %>% datatable() %>%
formatRound(columns = cols,  digits=3)
```

----

### When should you write a function?
You should consider writing a function whenever you’ve copied and pasted a block of code more than twice

```{r}
a <- rnorm(10)
b <- rnorm(10)
c <- rnorm(10)
d <- rnorm(10)
```

-----

### Without functions 
- introduce NA's 
```{r}
a - min(a)/(max(a) - min(a))
b - min(b)/(max(b) - min(b))
c - min(c)/(max(c) - min(c))
d - min(d)/(max(d) - min(d))
```

-----

### Function to work on this

- can introduce many features applied to all

```{r}
zero_one <- function(x, na_rm = T, round_digits = 3){
  min_x = min(x, na.rm = na_rm)
  max_x = max(x, na.rm = na_rm)
  diff_x = max_x - min_x
  ans = (x - min_x)/diff_x
  round(ans, digits = round_digits)
}
```

-----

### With Functions

```{r}
zero_one(a)
zero_one(b)
zero_one(c)
zero_one(d)
```

### Return value

```{r}
plus_one <- function(x){
  x = x+1
}
plus_one(5)
```

-----

```{r}
plus_one_return <- function(x){
  x = x+1
  x
}
plus_one_return(5)
```


```{r}
plus_one_return <- function(x){
  x = x+1
  return(x)
}
plus_one_return(5)
```

----

### Inputs to functions

- Most functions require some sort of input to determine what to compute.
The inputs to functions are called arguments. You specify them inside the
parentheses after the word "function."
- In the above a, b, c, d are the arguments

### Multiple inputs to functions
 - If has more than one argument, list them in the function 
signature, separated by commas.

```{r}

cor(x = iris$Sepal.Length,y= iris$Petal.Length)

```


-----

```{r}
fx <- function(x){
  x = x+1
  cat("x = ", x)
  x
}
```

```{r}

```

### Numeric defaults
- add cutpoint = 6

```{r}
number_greater <- function(age, cutpoint){
  age_cut <- age>=cutpoint
  no_greater <- sum(age_cut)
  return(no_greater)
  
}

age <- rnorm(100, mean = 6, sd = 3)
# if you dont specify cutpoint it will throw an error

number_greater(age = age, cutpoint = 6)
```


```{r}
number_greater <- function(age, cutpoint = 6){
  age_cut <- age>=cutpoint
  no_greater <- sum(age_cut)
  return(no_greater)
  
}

number_greater(age = age)
number_greater(age = age, cutpoint = 7)
```

-----

```{r}

x = 3
z = fx(x)
cat('z =', z)
cat('x =', x)

```

-----


### Functions inputs & outputs various R functions

### Assignments


```{r}

```

