---
title: "Decrementing Functions & Iterartion"
date: "5/24/2022"
output:
  revealjs::revealjs_presentation:
     theme: sky
     highlight: pygments
     transition: convex
  slidy_presentation: default
  html_document: default
  ioslides_presentation: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE)
```


```{r}
library(tidyverse)
library(DT)
```

## Iteration reducing duplication

- Remember our condition 

- For loop

```{r}

for (i in 1:5) {
  print(i)
}
```

-----


```{r}
df <- data.frame(
  a = rnorm(5),
  b = rnorm(5),
  c = rnorm(5),
  d = rnorm(5)
)
datatable(df)
```

-----

```{r}
median(df$a)
#> [1] -0.2457625
median(df$b)
#> [1] -0.2873072
median(df$c)
#> [1] -0.05669771
median(df$d)
#> [1] 0.1442633
```

-----

```{r}
output <- vector("double", ncol(df))  # 1. output
for (i in seq_along(df)) {            # 2. sequence
  output[[i]] <- median(df[[i]])      # 3. body
}
output
```
------


- Unknown sequence length (While loop)
Sometimes you don’t even know how long the input sequence should run for. This is common when doing simulations. For example, you might want to loop until you get three heads in a row. You can’t do that sort of iteration with the for loop. Instead, you can use a while loop. A while loop is simpler than for loop because it only has two components, a condition and a body

-----

- mostly used in machine learning

```{r}
output_while <- vector("double", ncol(df)) 
i =1
n_coldf = ncol(df)
while(i <= n_coldf){
  
   output_while[[i]] <- median(df[[i]])
   i = i+1 # decrementing function ensures that at the end the loop will terminate
   #print(i)
 
}

output_while
```

### Repeate Loop

- Use when you want to repeat a calculation until a condition is reached

- Mostly I recommend while  loop

```{r}
result <- 1
i <- 1
  
# test expression 
repeat {
  
   print(result)
     
   # update expression 
   i <- i + 1
   result = result + 1
  
   # Breaking condition
   if(i > 5) {
      break
   }
}
```


### Real life implementation 

- Implementation here in KEMRI by a friend

```{r}
n = 50
df <- data.frame(
  a = rnorm(n),
  b = rnorm(n),
  c = rnorm(n),
  d = rnorm(n)
)

```

-----

```{r}
miss_vars_perc <- seq(.5, .8, by = 0.01)
miss_vars_perc
miss_vars_perc <- sample(miss_vars_perc, ncol(df))
miss_vars_perc
```

----

```{r}
#output <- vector("double", ncol(df)) 
n_row  <- nrow(df)
for (i in seq_along(df)) { 
  n_rows_i = miss_vars_perc[i] *n_row
  n_rows_i = round(n_rows_i)
  number_rows = sample(n_row, n_rows_i)
  df[[i]][number_rows]=NA
}
```

-----

```{r}
datatable(df)
```

