---
title: "CORE ELEMENTS OF A SCRIPT"
author: "Mburu"
date: "6/1/2022"
output:
  revealjs::revealjs_presentation:
     theme: sky
     highlight: pygments
     transition: convex
  slidy_presentation: default
  html_document: default
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## data type
A data type, in programming, is a classification that specifies which type of value a variable has and what type of mathematical, relational or logical operations can be applied to it without causing an error. A string, for example, is a data type that is used to classify text and an integer is a data type used to classify whole numbers.

-----

## Common Data types in R

- character: "a", "swc"
- numeric: 2, 15.5
- integer: 2L (the L tells R to store this as an integer)
- logical: TRUE, FALSE
- complex: 1+4i (complex numbers with real and imaginary parts)

-----

## data structures
- Data Structures are a specialized means of organizing and storing data in computers in such a way that we can perform operations on the stored data more efficiently. Data structures have a wide and diverse scope of usage across the fields of Computer Science and Software Engineering.

## Common data structures in R

- atomic vector
- list
- matrix
- data frame
- factors

## Vectors

- Atomic vectors and lists  differ in terms of their elements’ types: for atomic vectors, all elements must have the same type; for lists, elements can have different types. While not a vector NULL is closely related to vectors and often serves the role of a generic zero length vector.

-----


### Atomic vectors

- There are four primary types of atomic vectors: logical, integer, double, and character (which contains strings)

----
```{r}

mybool <- c(TRUE, FALSE)#Boleen
myinteger <- c(1L, 2L, 3L) # integers
mydouble <- c(1.1, 3.2, 5.6) # floating # Real numbers
mycharacter <- c("moses", "a", "b", "c") # ccharacter

class(mybool)
class(myinteger)
class(mydouble)
class(mycharacter)
```

-----

- integer and double vectors are known as numeric
- Each of the four primary types has a special syntax to create an individual value, AKA a scalar

- You can determine the type of a vector with typeof()24 and its length with length()

### Missing values

- R represents missing, or unknown values, with special sentinel value: NA (short for not applicable). Missing values tend to be infectious: most computations involving a missing value will return another missing value

-----
```{r}
x <- c(3, 5, 4.1, 6, NA)
mean(x)
max(x)
sum(x)
sd(x)
```

-----

```{r}
NA > 5
10 * NA
!NA

```

-----

```{r}
NA ^ 0
NA | TRUE
NA & FALSE
```

-----

- Propagation of missingness leads to a common mistake when determining which values in a vector are missing:


```{r}
x <- c(NA, 5, NA, 10)
x == NA
is.na(x)
```

-----

### Testing and coercion

```{r}
is.logical(mybool)
is.integer(myinteger)
is.double(mydouble)
is.character(mycharacter)
```

-----
### Coercion

```{r}
as.numeric(mydouble)

myvec <- c(1, 2, 3., "a")
myvec
typeof(myvec)
as.numeric(myvec)
```

----

### Attributes
- You might have noticed that the set of atomic vectors does not include a number of important data structures like matrices, arrays, factors, or date-times. 

- names, a character vector giving each element a name.
- dim, short for dimensions, an integer vector, used to turn vectors into matrices or arrays

-----
```{r}
# When creating it: 
x <- c(a = 1, b = 2, c = 3)
x
# By assigning a character vector to names()
x <- 1:3
x
names(x) <- c("a", "b", "c")
x
# Inline, with setNames():
x <- setNames(1:3, c("a", "b", "c"))
x
```

-----

- Adding a dim attribute to a vector allows it to behave like a 2-dimensional matrix or a multi-dimensional array. Matrices and arrays are primarily mathematical and statistical tools, not programming tools

```{r}
mat <- matrix(1:4, ncol = 2)
mat
```

-----

### S3 atomic vectors
One of the most important vector attributes is class, which underlies the S3 object system. Having a class attribute turns an object into an S3 object, which means it will behave differently from a regular vector when passed to a generic function. Every S3 object is built on top of a base type, and often stores additional information in other attributes. 

-----

- Categorical data, where values come from a fixed set of levels recorded in factor vectors.

- Dates (with day resolution), which are recorded in Date vectors.

- Date-times (with second or sub-second resolution), which are stored in POSIXct vectors.

- Durations, which are stored in difftime vectors.

-----

```{r}
sex <- c("female", "female", "female", "female", "male", "male", "male", 
         "male", "male", "male", "male", "male", "male", "male", "male", 
         "female", "male", "male", "female", "female")
sex_factor <- factor(sex)

sex_factor

sex_num <- c(2, 2, 1, 1, 2, 1, 2, 1, 1, 
             1, 2, 2, 1, 1, 2, 1, 1, 2, 2, 2)

sex_num_factor <- factor(sex_num, 
                         levels = c(1, 2), 
                         labels = c("female", "male"))
sex_num_factor
```
-----



```{r}
sex_num_factor
typeof(sex_num_factor)
attributes(sex_num_factor)
```

-----

- Ordered factors are a minor variation of factors. In general, they behave like regular factors, but the order of the levels is meaningful (low, medium, high) (a property that is automatically leveraged by some modelling and visualisation functions).

-----

```{r}
grade <- ordered(c("b", "b", "a", "c"), levels = c("c", "b", "a"))
grade
```

-----

### Dates

- Date vectors are built on top of double vectors. They have class “Date” and no other attributes:

```{r}
today = Sys.Date()
mydates_char <- c("2022-05-31", "2022-05-30", "2022-05-29", "2022-05-28", "2022-05-27", 
             "2022-05-26", "2022-05-25", "2022-05-24", "2022-05-23", "2022-05-22")
summary(mydates_char)

mydates <- as.Date(mydates_char)

summary(mydates)
```

-----


### Date-times
Base R28 provides two ways of storing date-time information, POSIXct, and POSIXlt. These are admittedly odd names: “POSIX” is short for Portable Operating System Interface, which is a family of cross-platform standards. “ct” stands for calendar time (the time_t type in C), and “lt” for local time (the struct tm type in C). Here we’ll focus on POSIXct, because it’s the simplest, is built on top of an atomic vector, and is most appropriate for use in data frames. POSIXct vectors are built on top of double vectors, where the value represents the number of seconds since 1970-01-01.

-----

```{r}
Sys.timezone()
mytime = Sys.time()
june1st12pm <- as.POSIXct("2022-06-01 12:00:00", tz = "Africa/Nairobi")
june1st12pm
mytime
```

-----

###  Math operations on Dates

```{r}
prev_day <- Sys.Date()-1
prev_day
foll_day <- Sys.Date()+ 1
foll_day
mean(mydates)


```


-----

```{r}
## time difference
difftime(foll_day, prev_day, units = "hours")
difftime(foll_day, prev_day, units = "days")
```

-----

### strftime

- [Different formats representing dates](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/strptime)


----

## Subsetting

### Atomic Vectors

-----

- Positive integers return elements at the specified positions:
- In R the first position is 
```{r}
x <- c(2.1, 4.2, 3.3, 5.4)
x[1]
x[c(2, 4)]
```



-----

- Negative integers exclude elements at the specified positions:

```{r}
x[- 1]
x[- c(2, 4)]
```

-----

- Logical vectors select elements where the corresponding logical value is TRUE. This is probably the most useful type of subsetting because you can write an expression that uses a logical vector:

- This is the most common type of subsetting

-----

```{r}
y <- c(T, F, T, F)
y
x[y]
x[c(1, 3)]
x>2
x[x>2]
```

-----

### Lists

-----


```{r}
l1 <- list(
  1:3, 
  "a", 
  c(TRUE, FALSE, TRUE), 
  c(2.3, 5.9),
  list(c(4, 5, 6), "abcd")
)


```
-----


```{r}
typeof(l1)

l1
```


-----

#### Subset lists

```{r}
l1[[1]]
```

-----

```{r}
l2 <- list(country = c("Uganda", "Bangladesh", "Malawi"),
           collaborators = c("Chris", "Sayeem", "Chikondi"),
           sites = c("Kampala", "Dhaka", "Blantrye"))
```

-----

```{r}
l2
```

-----

```{r}

l2$country
l2$collaborators
l2$sites
```

-----

```{r}
l2[["country"]]
l2[["collaborators"]]
l2[["sites"]]
```


-----

### Data frames 

- Think of this as tabular data

```{r}
df <- data.frame(a = 1:2, b = 1:2)
df[1, ] # first row
df[, 1] # first col
```

